\setlength{\parindent}{0pt}
\chapter*{\vspace{-4\baselineskip} \bf ACKNOWLEDGEMENTS}

First of all, I would like to express the deepest appreciation to my thesis advisor Dr. Ogün Adebali. Dr. Ogün Adebali's meticulous comments were an enormous help to me, and without his immense support, patience, and encouragement, this thesis would not have materialized. I will always be grateful for having a chance to study in his lab, where my scientific background, technical knowledge, and skeptical thinking improved considerably. I also want to thank the rest of my thesis jury, Prof. Dr. Batu Erman, and Prof. Dr. Halil Kavaklı for their time and interest in my thesis project. 

I would like to thank all the members of ADEBALİLAB; Arda Çetin, Aylin Bircan, Berkay Selçuk, Burak İşlek, and Sezgi Kaya for their help, scientific comments, and especially for their friendship. Whether I need a company while sipping my coffee or I need a piece of insightful advice for my project, they have always been there for me, and I am grateful for that. I am also thankful to my undergraduate students Berk Turhan, Defne Çirci, and Zeynep Kılınç for their enthusiasm and friendship.

I would like to offer my special thanks to all my friends, who never stopped supporting me and relieving my mind when I needed them the most. Without them, my sanity would be at stake. My deepest gratitude goes to my family; my dad, who always trusted me, my mom with her constant care and interest in my studies. If it weren't for the importance she attached to education, I might not be a researcher today. To my grandmother, who is still tracking whether I finished my homework (thesis) or not, and to my brother Nedim Azgari, who without a doubt put the most effort to develop my personality and to boost my interest in learning. I have always admired his enthusiasm to improve and his mindset and taken him as a role model. Lastly, I would like to thank my fiancée Ecem Ornadis, who I considered being my greatest accomplishment. Since I knew her, she stands by my side, never letting me give up or back down. Without her support and trust, I might not even be a Master's student at Sabancı University in the first place.    

\clearpage\pagebreak